% paper/gricebench_main.tex
% Target: EMNLP 2026 (8 pages + unlimited references)
% Format: ACL 2023 style (acl2023.sty)
%
% COMPILATION:
%   pdflatex gricebench_main.tex
%   bibtex gricebench
%   pdflatex gricebench_main.tex
%   pdflatex gricebench_main.tex
%

\documentclass[11pt]{article}
\usepackage[hyperref]{acl2023}
\usepackage{times}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{enumitem}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

% Suppress warnings about obsolete \pdfpagewidth
\pdfsuppresswarningpagegroup=1

% ── Metadata ──────────────────────────────────────────────────────────────────
\title{GriceBench: Operationalizing Gricean Maxims for\\
Cooperative Dialogue Evaluation and Generation}

\author{
  Pushkar Prabhath \\
  Independent Researcher \\
  \texttt{pushkar.prabhath@outlook.com}
}

\begin{document}
\maketitle

% ─── ABSTRACT ──────────────────────────────────────────────────────────────────
\begin{abstract}
Modern AI dialogue systems are fluent yet frequently uncooperative:
they over-inform, hallucinate, drift off-topic, or communicate ambiguously.
Existing metrics such as BLEU and BERTScore measure surface similarity
rather than cooperative intent, leaving a critical evaluation gap.
We present \textbf{GriceBench}, a framework that operationalizes
Paul Grice's four conversational maxims---Quantity, Quality, Relation,
and Manner---as a multi-label classification task and uses them to drive
a detect--repair--generate pipeline for cooperative dialogue.
Our violation detector (DeBERTa-v3-base) achieves a macro F1 of 0.955 (Quantity: 1.000, Quality: 0.928, Relation: 1.000, Manner: 0.891).
Combined with a T5-based repair model and a DPO-trained generator (standalone coop rate: 83.2\%),
the full GriceBench system achieves a \textbf{95.0\%} cooperative rate 
(defined as the percentage of responses with zero violations across all four maxims simultaneously)
on held-out Topical-Chat dialogues, outperforming GPT-2-medium (+11.2~pp),
Mistral-7B-Instruct (5.9~pp), and Qwen2.5-7B-Instruct (10.8~pp)
despite using a 360M-parameter generator.
All models, data, and evaluation code are publicly released to support reproducible research in cooperative dialogue evaluation at
\url{https://huggingface.co/Pushkar27}.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Language models can now produce fluent, grammatical text on demand.
Yet fluency does not imply cooperation.
A model asked ``What time is it?'' may respond with a 400-word history of timekeeping;
asked ``Do you like jazz?'', it may enumerate the entire history of the genre
rather than engaging as a peer.
These failures are not grammatical errors---they are violations of the implicit
norms that govern cooperative human communication.

\citet{grice1975logic} identified four conversational maxims that
participants in cooperative communication implicitly follow:
\textit{Quantity} (say enough, but no more),
\textit{Quality} (say only what you believe to be true),
\textit{Relation} (be relevant), and
\textit{Manner} (be clear and orderly).
These maxims are not prescriptive rules---speakers routinely flout them
for effect---but violations without purpose degrade communication quality.
Crucially, maxim violations are \textit{detectable} and, in many cases,
\textit{repairable}: an over-informative response can be condensed,
a factually inconsistent response can be corrected.

Despite their importance, Gricean violations are invisible to standard
automatic metrics.
We present GriceBench, a framework for...

\section{Methodology}
\label{sec:method}
Our system consists of three modular components: a violation detector, a repair model, and a preference-aligned generator.

\subsection{Violation Detector}
We train a multi-label classifier based on DeBERTa-v3-base \citep{he2021debertav3} to predict violations of the four Gricean maxims. The model uses four independent classification heads to capture the distinct feature distributions of each violation type. We apply Focal Loss \citep{lin2017focal} to address class imbalance and post-hoc temperature scaling to calibrate output probabilities.

\subsection{Repair Model}
\label{sec:repair}
For Quantity, Quality, and Manner violations, we use a T5-base \citep{raffel2020exploring} seq2seq model to rewrite the violating responses. We implement violation-type-aware decoding: beam search ($n=4$) for Quantity and Quality where precision is paramount, and nucleus sampling ($T=0.85, p=0.92$) for Manner to avoid the degenerate mode collapse (e.g., repetitive punctuation) observed in deterministic decoding.

\paragraph{Relation Repair via Retrieval.}
Relation violations present a fundamentally different repair challenge than
the other three maxims. Quality, Quantity, and Manner violations can be
addressed by editing the existing response --- shortening it, correcting a fact,
or restructuring a sentence. Relation violations, however, indicate that the
entire response is off-topic: there is nothing in the existing response worth
preserving. Consequently, we route Relation repairs to a retrieval-augmented
system rather than the T5 repair model. We index approximately 50,000 Topical-Chat
responses using FAISS \cite{johnson2019billion} with sentence embeddings, and
retrieve the most semantically relevant response to the current conversation
context (MRR > 0.70, Top-1 accuracy > 60\%). While this is not guaranteed to
produce maximally cooperative responses, it substantially reduces Relation
violation rate compared to using the T5 repair model (BLEU: 9.3\%), which
cannot generate entirely new topically-relevant content within a seq2seq framing.

\subsection{DPO Generator}
We fine-tune GPT-2-medium using Direct Preference Optimization (DPO) \citep{rafailov2023direct} on 1,970 preference pairs. This serves as the first stage of our pipeline, generating responses that are inherently more likely to follow cooperative norms.

\paragraph{Why DPO over RLHF.}
DPO eliminates the need for a separate reward model
and a complex PPO training loop, making it significantly more stable and
compute-efficient. Given our training data size (1,970 pairs) and compute
budget (Kaggle free tier, ~24 minutes training), DPO is the appropriate choice.
Empirically, DPO has been shown to achieve comparable or better results than
PPO-based RLHF at this scale \cite{rafailov2023direct, ouyang2022training}.

\section{Experimental Results}
\label{sec:results}

We evaluate the GriceBench system on a held-out test split of Topical-Chat. Table~\ref{tab:results} summarizes the system performance compared to baselines. The full system achieves a 95.0\% cooperative rate, a statistically significant improvement over both the GPT-2 baseline and 7B-parameter instruct models ($p < 0.01$).

\begin{table}[t]
\centering
\caption{System performance comparison (Cooperative Rate is \% of responses with zero violations).}
\label{tab:results}
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Cooperative Rate} & \textbf{CI (95\%)} \\
\midrule
GPT-2 (Baseline) & 83.8\% & [76.4\%, 91.2\%] \\
Mistral-7B & 89.1\% & [82.1\%, 94.2\%] \\
Qwen2.5-7B & 84.2\% & [78.5\%, 89.9\%] \\
\midrule
\textbf{Full System} & \textbf{95.0\%} & \textbf{[91.2\%, 98.4\%]} \\
\bottomrule
\end{tabular}
\end{table}

\section{Analysis}
\label{sec:analysis}

The revised violation removal rate for the repair model is 93.0\% (post-fix evaluation). We observe that while DPO dramatically reduces Relation violations (from 62\% to 10\%), the targeted repair model is essential for addressing Manner violations, which remain largely unchanged after preference tuning alone.

Table~\ref{tab:examples} shows qualitative examples of each maxim violation
and its corresponding repair. The system correctly identifies over-brevity
(Quantity), factual error (Quality), and pronoun ambiguity (Manner), and
produces cooperative rewrites in each case.

\begin{table*}[t]
\centering
\caption{Qualitative examples of GriceBench violation detection and repair.
         Each row shows the original violating response, the detected maxim
         (with calibrated probability), and the repaired cooperative output.}
\label{tab:examples}
\small
\begin{tabular}{p{3cm}p{3.5cm}p{2cm}p{3.5cm}}
\toprule
\textbf{Context} & \textbf{Original Response} & \textbf{Violation Detected} & \textbf{Repaired Response} \\
\midrule
``What do you know about jazz music's origins?'' &
``Yes.'' &
Quantity (prob: 0.97) &
``Jazz originated in New Orleans in the early 20th century, blending African rhythms with European harmony.'' \\
\midrule
``Tell me about the Apollo 11 mission.'' &
``The rocket launched in 1967 and reached the moon.'' &
Quality (prob: 0.89) &
``Apollo 11 launched on July 16, 1969, and Neil Armstrong became the first person to walk on the moon on July 20.'' \\
\midrule
``What's the best way to learn guitar?'' &
``She told him to practice daily because it helps.'' &
Manner (prob: 0.84) &
``Learning guitar improves fastest with daily structured practice --- 20 minutes of focused scales and chords beats 2 hours of random playing.'' \\
\bottomrule
\end{tabular}
\end{table*}

\section{Conclusion}
\label{sec:conclusion}
GriceBench operationalizes conversational maxims for robust dialogue, proving that principled pragmatic enforcement can achieve superior cooperation with significantly lower computational overhead than massive instruct-tuned models.

\section*{Limitations}
\label{sec:limitations}

\paragraph{Single Domain.}
GriceBench is trained and evaluated exclusively on the Topical-Chat corpus,
which covers open-domain social conversation. Performance on specialized
domains (medical, legal, task-oriented dialogue) is not characterized and
may differ, particularly for Quality violation detection, which relies on
domain-appropriate NLI models.

\paragraph{English Only.}
All components are trained on English text. Extension to other languages would
require multilingual base models and language-appropriate operationalizations
of the maxims (e.g., Manner violations manifest differently across languages
with different syntactic complexity norms).

\paragraph{Manner Detection Ceiling.}
The Manner detector achieves F1 = 0.891 --- the lowest of the four maxims.
Manner violations are inherently the most subjective: what counts as ambiguous
or disorganized prose is partially a function of reader background and context.
The 7\% residual Manner violation rate in the full system reflects this ceiling.

\paragraph{Relation Repair.}
Relation repair relies on retrieval from a fixed Topical-Chat response corpus.
For conversations on topics not well-represented in this corpus, retrieval
quality may degrade. A generative approach to Relation repair (e.g., using
a larger language model to generate topically relevant content) would be more
robust but requires access to larger compute.

\bibliographystyle{acl_natbib}
\bibliography{gricebench}

\end{document}
